Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-07-27 Report/Sections/intro.tex
Update Intro- Nancy

Movement data are sometimes modelled as a set of curves that can  be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).  In \cite{Morris:2007}, a child's activity level throughout the day can be viewed as a curve, specifically, as activity as a function of time.   
In the data  we consider, a single dive of a whale can be viewed as a curve:  acceleration as a function of time.
FDA was originally developed for curves that can be considered independent replicates.
However, the structure of the data is often more complex.

Our goal is to develop movement analysis methods that provide subject-area insight  into the complex structure of curve data.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure and dependence from dive to dive via a hidden Markov model (HMM).
We model within dive variation by combining an HMM and a time series.
Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
While our methodology is generally applicable, we  showcase our methods in a  study of the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive,  behaviour that seems to indicate some fine-scale short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

Particular types of between curve dependence structures have received attention in the FDA literature, 
notablly multilevel models that incorporate dependence via random effects.
For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
Perhaps the first work with this multilevel structure is \cite{Bromback:1998} in a study of ....   More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012}} and Crainiceanu [REF]. 
For our data, the multilevel models are not appropriate, since they do not account for the temporal dependence.

FDA researchers have modelled temporal dependence in a sequence of curves  via functional time series, extending the ideas of classic time series to model how one curve evolves into the next curve.  See, for instance, \cite{Kokoszka:2018}. 
While  functional time series methods can be applied to our dive data, they would not account for the different dive types.
Our HMM models account for both temporal dive-to-dive dependence and dive types.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.   Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017,Mcclintock:2020}, where
one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
However, HMMs have seen little use in FDA.  
One exception is the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here. 
Part of the purpose of our article is to introduce the FDA community to these very useful approaches to modelling complex curves.

@@

We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.

>>>>

Movement data are sometimes modelled as a set of curves that can  be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).  In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve, specifically, as activity as a function of time.   
In the data  we consider, a single dive of a whale is viewed as several curves:  acceleration and depth as  functions of time.
FDA was originally developed for curves that can be considered independent replicates.
However, the structure of the data is often more complex.

Our goal is to develop movement analysis methods that provide subject-area insight  into the complex structure of curve data.
We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
combining existing methods in novel ways.
We enforce temporal structure and dependence from dive to dive via a hidden Markov model (HMM).
We model within dive variation by combining an HMM and a time series.
Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
While our methodology is generally applicable, we  showcase our methods in a  study of the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.  
For more information about the data, see Section \ref{sec:case_study}.
Figure \ref{fig:data} shows the data we will analyze.
The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer,   
with axes oriented relative to the whale's forward movement.
Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
The top panel  shows a summary  of the acceleration, and from this we can see fine-scale behaviour  that seems to indicate some  short-term periodicity. 
We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.

Models for particular types of between curve dependence structures appear in the FDA literature, 
notably multilevel models that incorporate dependence via random effects.
For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
In a classic early paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.  More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF]. 
For our data, multilevel models are not appropriate, since they do not account for the temporal dependence in the sequence of dives nor do they account for different dive types.

FDA researchers have modelled temporal dependence in a sequence of curves  via functional time series, extending the ideas of classic time series to model how one curve evolves into the next curve.  See, for instance, \cite{Kokoszka:2018}. 
While  functional time series methods can be applied to our dive data, they would not account for the different dive types.
Our HMM models account for both temporal dive-to-dive dependence and different types of dives.

@@

From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
Alternatively, this structure is sometimes generated via random regression.
However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence. Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.   Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve. The distribution of this ``wiggliness'' depends on the hidden state.

Our work relies on work in the vast field of methodology for the analysis of animal movement
\citep{Hooten:2017,Mcclintock:2020}, where
one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
However, HMMs have seen little use in FDA.  
One exception is the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here. 
Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.

@@

We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
The CarHMM  explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
